Estimating epidemiological quantities from repeated cross-sectional prevalence measurements

### 2022-03-17 docs/paper.Rmd
minor paper edits; waning estimates now monthly (#33)

---
title: "Estimating epidemiological quantities from repeated cross-sectional prevalence measurements"
author: Sam Abbott, Sebastian Funk
bibliography: references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
date: "`r format(Sys.Date(), format = '%B %d, %Y')` - WORK IN PROGRESS DRAFT"
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
  bookdown::html_document2: default
---

@@

params_table <- here::here("outputs", "example-parameters.csv")
if (!file.exists(params_table)) {
  source(here::here("scripts", "simple-example.R"))
}
params <- fread(params_table)
params[name == "gamma",
       name := ifelse(index == 1, "gamma (infection)", "gamma (vaccination)")
]
params[,
  Description := fcase(
    name == "beta", "Proportion infected that seroconvert",
    name == "gamma (infection)", "Antibody waning following infection (daily rate)",
    name == "gamma (vaccination)", "Antibody waning following vaccination (daily rate)",
    name == "delta", "Proportion vaccinated that seroconvert",
    name == "k", "Efficacy adjustment of immunity following infection",
    name == "l", "Efficacy adjustment of immunity following vaccination")
]
biol_params <- params[!is.na(Description)]
biol_params[, 
  `Estimate (90% CI)` := paste0(
    signif(q50, 2), " (", signif(q5, 2), "--", signif(q95, 2), ")"
  )
]
biol_params <- biol_params[,
  .(Parameter = name, Description, `Estimate (90% CI)`)
]

@@

Our estimated parameters of antibody dynamics are averages across a various combinations of vaccine types and individual factors that are known to affect immunological responses to either infection or vaccination, particularly age [@ward2022]. 
We estimated that `r 100 - signif(params[name == "beta"]$median, 2) * 100`% (90% credible interval, CI: `r 100 - signif(params[name == "beta"]$q95, 2) * 100`--`r 100 - signif(params[name == "beta"]$q5, 2) * 100`) of individuals do not seroconvert after infection, compared to 24% estimated from the same study population, but also lower estimates such as 10% in a different study [@gudbjartsson2020]. 
We further estimated that `r signif(params[name == "delta"]$median, 2) * 100`% (90% CI: `r signif(params[name == "delta"]$q95, 2) * 100`--`r signif(params[name == "delta"]$q5, 2) * 100`) of individuals seroconverted following vaccination, in line with high such proportions estimated in healthcare workers [@eyre2021].
Our estimates of waning correspond to an average duration of immunity of `r signif(1 / params[name == "gamma (infection)"]$median, 2)` days (90% CI: `r signif(1 / params[name == "gamma (infection)"]$q95, 2)`--`r signif(1 / params[name == "gamma (infection)"]$q5, 2) `) following infection and `r signif(1 / params[name == "gamma (vaccination)"]$median, 2)` days (90% CI: `r signif(1 / params[name == "gamma (vaccination)"]$q95, 2)`--`r signif(1 / params[name == "gamma (vaccination)"]$q5, 2)`) following vaccination.

>>>>

---
title: "Estimating epidemiological quantities from repeated cross-sectional prevalence measurements"
author: Sam Abbott, Sebastian Funk
bibliography: references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
date: "`r format(Sys.Date(), format = '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
  bookdown::html_document2: default
---

@@

params_table <- here::here("outputs", "example-parameters.csv")
if (!file.exists(params_table)) {
  source(here::here("scripts", "simple-example.R"))
}
params <- fread(params_table)
params <- melt(params, id.vars = c("name", "index"))
params[name == "gamma", value := 1 - exp(-365.25 / 12 * value)]
params <- dcast(params, name + index ~ variable)
params[name == "gamma",
       name := ifelse(index == 1, "gamma (infection)", "gamma (vaccination)")
]
params[,
  Description := fcase(
    name == "beta", "Proportion infected that seroconvert",
    name ==
    "gamma (infection)", "Antibody waning following infection (per month)",
    name ==
    "gamma (vaccination)", "Antibody waning following vaccination (per month)",
    name == "delta", "Proportion vaccinated that seroconvert",
    name == "k", "Efficacy adjustment of immunity following infection",
    name == "l", "Efficacy adjustment of immunity following vaccination")
]
biol_params <- params[!is.na(Description)]
biol_params[, 
  `Estimate (90% CI)` := paste0(
    signif(q50, 2), " (", signif(q5, 2), "--", signif(q95, 2), ")"
  )
]
biol_params <- biol_params[,
  .(Parameter = name, Description, `Estimate (90% CI)`)
]

@@

Our estimated parameters of antibody dynamics are averages across a various combinations of vaccine types and individual factors that are known to affect immunological responses to either infection or vaccination, particularly age [@ward2022]. 
We estimated that `r 100 - signif(params[name == "beta"]$median, 2) * 100`% (90% credible interval, CI: `r 100 - signif(params[name == "beta"]$q95, 2) * 100`--`r 100 - signif(params[name == "beta"]$q5, 2) * 100`) of individuals do not seroconvert after infection, consistent with the 24% estimated from the same study population, but also lower estimates such as 10% in a different study [@gudbjartsson2020]. 
We further estimated that `r signif(params[name == "delta"]$median, 2) * 100`% (90% CI: `r signif(params[name == "delta"]$q95, 2) * 100`--`r signif(params[name == "delta"]$q5, 2) * 100`) of individuals seroconverted following vaccination, in line with high such proportions estimated in healthcare workers [@eyre2021].
Our estimates of waning suggest that detectable antibodies decrease by `r signif(params[name == "gamma (infection)"]$median, 2) * 100`% (90% CI: `r signif(params[name == "gamma (infection)"]$q5, 2) * 100`--`r signif(params[name == "gamma (infection)"]$q95, 2) * 100`) following infection and `r signif(params[name == "gamma (vaccination)"]$median, 2) * 100`% (90% CI: `r signif(params[name == "gamma (vaccination)"]$q5, 2) * 100`--`r signif(params[name == "gamma (vaccination)"]$q95, 2) * 100`) following vaccination.
All these values depend on the specific cutoff used for seropositivity and combine a range of vaccines, and they ignore additional effects from receiving multiple doses of vaccine, becoming infected as well as vaccinated.
They cannot be compared directly to estimates of vaccine efficacy or waning thereof.
